AA: The Algorithmic Autoregulation (Distributed Software Development) Methodology

### 2013-08-07 paper.tex
small changes in the point of the reviewer0

One of the defining features of modern times is the widening geographical
distribution of software teams~\cite{last2003} creating what is called Global
Software Development (GSD)~\cite{german2003,Fryer,Begel}.  
An example is the free software movement. Projects and institutions like Mozilla
Foundation has several employees and thousands of voluntary developers
distributed across many countries. The same is true for GNOME~\cite{german2003},
OpenBSD, MySQL or Apache Software Foundation, to cite just a 
few of the most active projects.\footnote{Ohloh, the open source network, has a more complete
and constantly updated list of the most active projects on-line at
\url{www.ohloh.net}.} Beyond the free and open software community, GSD has a growing
popularity in every niche of the software industry as a whole, even among those
distributing their software with proprietary licenses. This phenomenon is
attributed to a variety of factors such as a larger labor pool, natural
globalization of software companies and foundations or even the premise of
cheaper cost of production~\cite{komi2005}.

@@

A brief overview of current work on GSD methodologies related to AA is presented
in Section~\ref{related-work}. In Section~\ref{aa-methodology} the most relevant
characteristics of the AA methodology are outlined. In the same section, it
is reported an actual use case of AA for coordinating a team of 9 paid developers 
% nao era mais de 9?
during the second half of 2011, as well as a broader use case of AA from 2012 to
the present time.  In Section~\ref{conclusions}, there are overall conclusions
and indicative of future possibilities for the practical use of AA in other types of
teams of software developers or organizations working on non-software distributed
activities.

@@

Second, cultural distance encompasses differences on organizational and
natural culture. Spoken language, unit and ethnic values are common
forms of such distance. Some companies prefer to allocate development
units in foreign locations with minimal cultural distance (\eg, an
American company may prefer Ireland due to spoken language
similarity~\cite{carmel2001}). Third, the temporal distance that
hampers synchronous communications like telephone or
video conferences. Units of developers working on different time-zones
are concerned with managing of their agenda guided by this temporal
distance.

Targeting geographical distance, Carmel~\cite{carmel2001} suggests a
strategy to reduce intensive collaboration. His approach divides the
whole software life-cycle into levels of complexity. Each level has a
degree of collaboration. For example, some developers working on a
project with high collaboration level should use the follow-the-sun
approach: when concluding the work day, they pass their work to the
team working on another time-zone. Other tactics are suggested by the
same author to deal with the three distances, such as separating foreign
units of developers in time-zone bands.

Battin~\etal~\cite{battin2001} propose and discuss their experiments using
specific methodologies created for the distributed development centers from
Motorola (at the time having 25+ software development centers worldwide). These
methodologies included constant communication with critical units, incremental
integration and schedules based on time-zones distributed to developers on 6
countries from 3 continents.

In considering free software projects instead of companies,
similar factors are present and specialized methodologies
arise. German~\cite{german2003} provides a concise review of
the methodologies used by the GNOME project, one of the most active of all free
software projects. The manuscript is centered on the software architecture. It
begins by explaining that GNOME is separated into modules (76 on version 2.4, to be
precise) and each module has one maintainer who divides her modules into
separate parts in which other developers can work on independent tasks, along
other responsibilities. As in other free software projects, all 
development was done using a \emph{bugtracker} for bug and issue management,
mail lists and Internet Relay Chat (IRC) for discussion and communication and a
version control system like Git or Mercurial. Periodical (commonly yearly)
conferences like GUADEC is common on free and open source projects for
face-to-face meetings and is based on a different place each time.

@@

Some of the strategies for GSD mentioned on the previous section are based on
complex methodologies and many were built for a specific company or software
center. This section describes an alternative methodology based on a simple idea: small
working sessions logged by a computational tool. Figure~\ref{fig:mm} summarizes
this methodology.

@@

From the developer's perspective, the AA methodology is based on generating
small, high level reports called \emph{micrologs} or \emph{AA shouts} of what
they are doing in a specific period of time.  This timeframe can be
between 5 to 15 minutes in our proposed practice, depending on what is most
convenient for the developer and the team. An \emph{AA Session} is a focused
period of continuous work, lasting about 2 hours in our proposed regime, in
which the developer issues a collection of these short periodical
\emph{micrologs}. The developer can set reminders or alerts to show up when it
is time to \emph{microlog}. The objective of the flexible timeframe and alert
scheme is to \emph{minimize developer overhead} during his AA session.  In this
way, the developer can issue micrologs while staying maximally focused on
his code. Each microlog may be sent directly to an on-line server, or stored
locally in a temporarily database for sending/pushing later on.  This enables
offline micrologging and periodical alerting.

@@

Easy and effective GSD team management is the main purpose of the AA
methodology. The proposed methodology was applied to a group of 9 developers in
July of 2011, 3 of which are coauthors of the present work, on what was named Lab
Macambira.\footnote{LabMacambira.sf.net: \url{http://labmacambira.sf.net}.} The
main objective of the team was to work on an array of strategic free software
projects in the broad audiovisual and web categories, contributing directly to
their development, submitting bug patches or committing new features to their
source code.

@@

In one month, each developer officially contributed to one or many free software
projects. Many developers started the initiation training with no knowledge of what
is free software and ended that period becoming a free software developer.
During that month, the same team of trainees also developed the first version of the AA
system and used AA to manage their activities, even while developing
the other aforementioned free software projects. Thus, AA and the associated
software system was tested, prototyped, and developed in close contact with
actual practice.  The source code of AA --- both the client that sends the
logs and the AA Web server --- is public available as free software\footnote{AA source
  code:
  \url{http://labmacambira.git.sourceforge.net/git/gitweb.cgi?p=labmacambira/aa}.}
and all the actual AA log data of the entire team of Lab Macambira from 2011 to
the present time is also on-line\footnote{Logs of AA sessions:
  \url{http://labmacambira.git.sourceforge.net/git/gitweb.cgi?p=labmacambira/paainel}.}.

After the initial training period of 1 month, the initiated `Macambiras'
worked during 6 additional months on a large range of free software projects,
divided into work groups --- each work group focusing on a specific theme like
video, audio and web --- and funded by contracts, freelance, and support of the
Pont\~{a}o N\'{o}s Digitais\footnote{\url{http://nosdigitais.teia.org.br}.}.
Table~\ref{tabela:criados} has a list of \emph{new} free software applications
created by `Lab Macambira' since July 2011.

While using the AA system, developers learned to work asynchronously with others
and got used to the habit of periodically updating their status on their
projects. Each programmer was given the chance to work with considerable
freedom, in any place and time of preference. The strictest required
responsibility was that of using AA for at least one session of 2h per day,
while working on the agreed upon tasks.  The online pAAnel allowed each
developer to quickly grasp activities from others while avoiding interrupting them, a
process further aided by the screencasts. Adjustments to the task deadlines and
milestones (which were managed in Trac) were done based on verified progress of
individuals and labMacambira.sf.net team as a whole. The various newcomers
benefited from a fast learning team by the use of AA as a flexible and simple
transparency system. Updates from the team were not transmitted on a
person-to-person basis, but rather on a person-to-team basis based on the
available online progress information.

@@

AA is not limited to a work-management tool, but acts as \emph{a
methodology to improve the time sensibility of individuals}, helping divide their
complex tasks in time into small chunks or sessions, and also reducing the need
of extensive reports or unnecessary meetings. By asking users to write a minimal
text sentence as a continuous log feed, the proposed methodology avoids
disturbing the flow of developers which are heavily concentrated on programming:
developers just have to type a few words and go back to coding.

@@

The AA methodology is not restricted to software development, even though it was
designed for the latter. As of this writing there is an entertainment studio,
Pula Pirata\footnote{ \url{www.pulapirata.com}.}, that has been using AA to
manage their creative activities.  Other people with no software background,
like social scientists, musicians and activists also have been using AA and
contributing for its broader improvement.

>>>>

One of the defining features of modern times is the widening geographical
distribution of software teams~\cite{last2003} creating what is called Global
Software Development (GSD)~\cite{german2003,Fryer,Begel}.  
An example is the free software movement. Projects and institutions like Mozilla
Foundation has several employees and thousands of voluntary developers
distributed across many countries. The same is true for GNOME~\cite{german2003},
OpenBSD, MySQL or Apache Software Foundation, to cite just a 
few of the most active projects.Beyond the free and open software community, GSD has a growing
popularity in every niche of the software industry as a whole, even among those
distributing their software with proprietary licenses. This phenomenon is
attributed to a variety of factors such as a larger labor pool, natural
globalization of software companies and foundations or even the premise of
cheaper cost of production~\cite{komi2005}.\footnote{The open source network Ohloh has a more complete
and constantly updated list of the most active projects on-line at
\url{www.ohloh.net}.}

@@

A brief overview of current work in GSD methodologies related to AA is presented
in Section~\ref{related-work}. In Section~\ref{aa-methodology} the most relevant
characteristics of the AA methodology are outlined. In the same section, it
is reported an actual use case of AA for coordinating a team of 9 paid developers 
% nao era mais de 9?
during the second half of 2011, as well as a broader use case of AA from 2012 to
the present time.  In Section~\ref{conclusions}, there are overall conclusions
and indicative of future possibilities for the practical use of AA in other types of
teams of software developers or organizations working on non-software distributed
activities.

@@

Second, cultural distance encompasses differences in organizational and
natural culture. Spoken language, unit and ethnic values are common
forms of such distance. Some companies prefer to allocate development
units in foreign locations with minimal cultural distance (\eg, an
American company may prefer Ireland due to spoken language
similarity~\cite{carmel2001}). Third, the temporal distance that
hampers synchronous communications like telephone or
video conferences. Units of developers working on different time-zones
are concerned with managing of their agenda guided by this temporal
distance.

Targeting geographical distance, Carmel~\cite{carmel2001} suggests a
strategy to reduce intensive collaboration. His approach divides the
whole software life-cycle into levels of complexity. Each level has a
degree of collaboration. For example, some developers working on a
project with high collaboration level should use the follow-the-sun
approach: when concluding the work day, they pass their work to the
team working in another time-zone. Other tactics are suggested by the
same author to deal with the three distances, such as separating foreign
units of developers in time-zone bands.

Battin~\etal~\cite{battin2001} propose and discuss their experiments using
specific methodologies created for the distributed development centers from
Motorola (at the time having 25+ software development centers worldwide). These
methodologies included constant communication with critical units, incremental
integration and schedules based in time-zones distributed to developers on 6
countries from 3 continents.

In considering free software projects instead of companies,
similar factors are present and specialized methodologies
arise. German~\cite{german2003} provides a concise review of
the methodologies used by the GNOME project, one of the most active of all free
software projects. The manuscript is centered on the software architecture. It
begins by explaining that GNOME is separated into modules (76 on version 2.4, to be
precise) and each module has one maintainer who divides her modules into
separate parts in which other developers can work on independent tasks, along
other responsibilities. As in other free software projects, all 
development was done using a \emph{bugtracker} for bug and issue management,
mail lists and Internet Relay Chat (IRC) for discussion and communication and a
version control system like Git or Mercurial. Periodical (commonly yearly)
conferences like GUADEC is common on free and open source projects for
face-to-face meetings and is based in a different place each time.

@@

Some of the strategies for GSD mentioned in the previous section are based on
complex methodologies and many were built for a specific company or software
center. This section describes an alternative methodology based on a simple idea: small
working sessions logged by a computational tool. Figure~\ref{fig:mm} summarizes
this methodology.

@@

From the developer's perspective, the AA methodology is based on generating
small, high level reports called \emph{micrologs} or \emph{AA shouts} of what
they are doing in a specific period of time.  This timeframe can be
between 5 to 15 minutes in our proposed practice, depending on what is most
convenient for the developer and the team. An \emph{AA Session} is a focused
period of continuous work, lasting about 2 hours in our proposed regime, in
which the developer issues a collection of these short periodical
\emph{micrologs}. The developer can set reminders or alerts to show up when it
is time to \emph{microlog}. The objective of the flexible timeframe and alert
scheme is to \emph{minimize developer overhead} during his AA session.  In this
way, the developer can issue micrologs while staying maximally focused in
his code. Each microlog may be sent directly to an on-line server, or stored
locally in a temporarily database for sending/pushing later on.  This enables
offline micrologging and periodical alerting.

@@

Easy and effective GSD team management is the main purpose of the AA
methodology. The proposed methodology was applied to a group of 9 developers in
July of 2011, 3 of which are coauthors of the present work, on what was named Lab
Macambira.The main objective of the team was to work on an array of strategic free software
projects in the broad audiovisual and web categories, contributing directly to
their development, submitting bug patches or committing new features to their
source code.\footnote{LabMacambira.sf.net: \url{http://labmacambira.sf.net}.}

@@

In one month, each developer officially contributed to one or many free software
projects. Many developers started the initiation training with no knowledge of what
is free software and ended that period becoming a free software developer.
During that month, the same team of trainees also developed the first version of the AA
system and used AA to manage their activities, even while developing
the other aforementioned free software projects. Thus, AA and the associated
software system was tested, prototyped, and developed in close contact with
actual practice.  The source code of AA --- both the client that sends the
logs and the AA Web server --- is public available as free software and all the actual AA log data of the entire team of Lab Macambira from 2011 to
the present time is also on-line.\footnote{AA source
  code:
  \url{http://labmacambira.git.sourceforge.net/git/gitweb.cgi?p=labmacambira/aa}.} \footnote{Logs of AA sessions:
  \url{http://labmacambira.git.sourceforge.net/git/gitweb.cgi?p=labmacambira/paainel}.}

After the initial training period of 1 month, the initiated `Macambiras'
worked during 6 additional months on a large range of free software projects,
divided into work groups --- each work group focusing on a specific theme like
video, audio and web --- and funded by contracts, freelance, and support of the
Pont\~{a}o N\'{o}s Digitais.
Table~\ref{tabela:criados} has a list of \emph{new} free software applications
created by `Lab Macambira' since July 2011.\footnote{\url{http://nosdigitais.teia.org.br}.}

While using the AA system, developers learned to work asynchronously with others
and got used to the habit of periodically updating their status on their
projects. Each programmer was given the chance to work with considerable
freedom, in any place and time of preference. The strictest required
responsibility was that of using AA for at least one session of 2h per day,
while working on the agreed tasks.  The online pAAnel allowed each
developer to quickly grasp activities from others while avoiding interrupting them, a
process further aided by the screencasts. Adjustments to the task deadlines and
milestones (which were managed in Trac) were done based on verified progress of
individuals and labMacambira.sf.net team as a whole. The various newcomers
benefited from a fast learning team by the use of AA as a flexible and simple
transparency system. Updates from the team were not transmitted in a
person-to-person basis, but rather in a person-to-team basis based on the
available online progress information.

@@

AA is not limited to a work-management tool, but acts as \emph{a
methodology to improve the time sensibility of individuals}, helping divide their
complex tasks in time into small chunks or sessions, and also reducing the need
of extensive reports or unnecessary meetings. By asking users to write a minimal
text sentence as a continuous log feed, the proposed methodology avoids
disturbing the flow of developers which are heavily concentrated in programming:
developers just have to type a few words and go back to coding.

@@

The AA methodology is not restricted to software development, even though it was
designed for the latter. As of this writing there is an entertainment studio,
Pula Pirata, that has been using AA to manage their creative activities.  Other people with no software background,
like social scientists, musicians and activists also have been using AA and
contributing for its broader improvement.\footnote{ \url{www.pulapirata.com}.}
